---
title: "Investigating the patterns of damage from various weather related natural disasters"
author: "Jimi Damon"
date: "06/21/2015"
output: html_document
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='./figure/',
                     warning=FALSE, message=FALSE)


library(dplyr)
library(ggplot2)
library(lubridate)
```


# Syopsis
With global scientific consensus corroborating the direct causal relationship between mankind's consumption of fossil fuels and global climate change, it is ever more important that Government resources be allocated to handle the new weather related disasters that are spawned by this changes of the Earth's climate.  In this dossier , we will be investigating ways to assess which weather related 

This analyis will involve using the storm data provided by the national weather service between the years of 1950 and 2011. The assumptions for this analyis based on the years of coverage 


```{r,echo=TRUE}
dat <- read.csv("stormdata.csv")
```


# Data Processing

Something goes in here

Will be investigating which types of events are most harmful to human health

Which events have the largest economic consequences


# Results

Something goes in here too

# Conclusion 

Something here too
